Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge

### 2025-02-10 paper/main.tex
Udpates

The \textsc{ExpertInLoop} algorithm adds a new edge in each iteration. The edge
to be added is selected using the \textsc{Expand} algorithm, which returns a
potential edge at random based on the iteration order in the algorithm.
However, if the residual association between the selected variables is low,
adding the edge may result in only a marginal improvement in the model's
overall fit. As the algorithm requires manual intervention in each iteration in
terms of domain knowledge, it would be beneficial to prioritize the edges that
improve the model's fit the most.

Therefore, an alternative to selecting an edge at random is to rank potential
edges based on the residual association between the variables given the DAG.
For quantifying the residual association between variables, we can use the
effect size measures of the CI test. So, for a d-separation query $ D_G(X, Y,
\mathbf{Z}) $, we can quantify the residual association between $ X $ and $ Y $
conditioned on $ \mathbf{Z} $, $ \phi(X, Y, \bm{Z}) $ as the effect size of the
CI test $ X \ci Y | \mathbf{Z} $. A greedy approach could be to choose the edge
that has the highest absolute residual association between the variables.
Algorithm~\ref{algo:ranked_expand} outlines this approach. This approach has
the benefit that it prioritizes edges that improve the model fit the most. We use
the \textsc{RankedExpand} in all of our experiments in the next section.

@@

The \textsc{ExpertInLoop} algorithm requires two oracles, one that can answer
d-separation queries and an ancestral oracle that can give the orientation or
non-existence of edges between pairs of variables. For answering the
d-separation queries, we use a CI test, the partial correlation test with a
p-value threshold of $ 0.05 $. Secondly, to tell the direction between edges or
non-existence of edges, we simulate an expert with accuracy $ \alpha $ as
follows:

\begin{equation}
	\begin{split}
		x &= \textrm{rand}([0, 1]) \\
		E(\alpha) &= \begin{cases} 
			\mathrm{TrueDir}(X, Y),  & \textrm{if  } x <= \alpha \\
			\textrm{rand}(X \rightarrow Y, Y \leftarrow X, \textrm{None}) & \textrm{otherwise} \\
				\end{cases} \\
	\end{split}
\end{equation}

@@

\begin{equation}
	\begin{split}
	\mathrm{TrueDir}(X, Y) &= \begin{cases}
					X \rightarrow Y, & \textrm{if } X \rightarrow Y \in G \\
					Y \rightarrow X, & \textrm{if } Y \rightarrow X \in G \\
					\textrm{None}, & \textrm{otherwise }
				  \end{cases}
	\end{split}
\end{equation}

An important point to note here is that the effective accuracy of $ E $ is
higher than $ \alpha $ as even when giving a random answer, there is a $ 1 $ in
$ 3 $ chance that the answer is correct. The effective accuracy is:

@@

We analyzed the performance of \textsc{ExpertInLoop} algorithm by comparing it
with three other automated algorithms: PC, Hill-Climb Search, and Greedy
Equivalence Search (GES). For the analysis, we simulated $ 500 $ samples from
known DAGs and use the algorithms to recover the original DAG. We start by
generating a random DAG on $10$ nodes and use linear models with random effects
to generate the data. We vary the edge probability of the DAG to generate DAGs
with varying density. We repeat the data generation $30$ times for each density
value for the DAG. We compare the learned DAG to the original DAG using two
metrics: Structural Hamming Distance (SHD) and Structural Intervention Distance
(SID) \citep{Peters2015}.

\begin{equation}
	Y = \bm{\beta} \mathrm{Pa}_G(Y) + \mathcal{N}(0, 1)
\end{equation}

As the automated algorithms can only recover the CPDAG from data, and our
method recovers the DAG. To compare them fairly, we computed the SHD and SID
for all possible orientations of the CPDAG to show the best and worst case
scenarios. The results are shown in Figure~\ref{fig:shd_sid}. For SHD, the
performance of \textsc{ExpertInLoop} is comparable to the automated algorithms
for $ \alpha = 0.3 (\alpha_{\textrm{eff}} = 0.53 $, and performs better for
higher $ \alpha $ values. For SID, the \textsc{ExpertInLoop} performs better in
denser DAGs.

% \begin{verbatim}
% 	Input: Dataset $ D $ on variables $ \bm{X} $
% 	Output: Learned DAG G.
% 	
% 	function compute_effects(G):
% 		pvalues <- dict()
% 		effects <- dict()
% 		for X_i, X_j in X:
% 			if edge between X_i, X_j:
% 				p_values[(X_i, X_j)] <- pvalue(X_i, X_j, pa_{\bar{G}}(X_i) \cup pa_{\bar{G}}(X_j))
% 				effects[(X_i, X_j)] <- \phi(X_i, X_j, pa_{\bar{G}}(X_i) \cup pa_{\bar{G}}(X_j))
% 			else:
% 				p_values[(X_i, X_j)] <- pvalue(X_i, X_j, pa_{\bar{G}}(X_i) \cup pa_{\bar{G}}(X_j))
% 				effects[(X_i, X_j)] <- \phi(X_i, X_j, pa_{\bar{G}}(X_i) \cup pa_{\bar{G}}(X_j))
% 		return p_values, effects
% 	
% 	$ G $ <- Empty Graph
% 	p_values, effects <- compute_effects(G)
% 	while (p_values > p_thres, 
% \end{verbatim}
% 
% \begin{algorithm}
% 	\KwData{Data set $ D $ on variables $ \bm{X} $, pthres, effectthres}
% 	\KwResult{The learned DAG: $ G $}
% 	\BlankLine
% 	$ G \leftarrow $ Empty Graph
% \end{algorithm}

>>>>

The \textsc{ExpertInLoop} algorithm adds a new edge in each iteration. The edge
to be added is selected using the \textsc{Expand} algorithm, which returns a
potential edge at random based on the iteration order. However, if the residual
association between the selected variables is low, adding the edge may result
in only a marginal improvement in the overall model fit. Given that the
algorithm requires expert intervention in each iteration to specify the edge
orientation, it is beneficial to prioritize edges that contribute the most to
improving the model.

To achieve this, instead of selecting edges randomly, we propose ranking
potential edges based on their residual association given the current DAG. This
residual association can be quantified using the effect size measures from the
CI tests used for querying the d-separation in Algorithm~\ref{algo:expand}.
Specifically, for a d-separation query $ D_G(X, Y, \mathbf{Z}) $, we quantify
the residual association, $ \phi(X, Y, \bm{Z}) $, as the effect size of the CI
test $ X \ci Y \rvert \bm{Z} $. In Algorithm~\ref{algo:ranked_expand}, we
outline a greedy approach to edge selection using the effect size measure. The
algorithm selects the edge that has the highest absolute residual association,
resulting in prioritization of edges that contribute the most to improving the
model. We use this \textsc{RankedExpand} in all of our experiments in the
following sections.

@@

The \textsc{ExpertInLoop} algorithm requires two oracles, one that can answer
d-separation queries and an ancestral oracle that can give the causal direction
between variable pairs. For the experiments, we simulate the d-separation
oracle using a CI test with a p-value threshold of $ 0.05 $. To simulate the
edge orientation oracle, we simulate a domain expert that can give the correct
causal relationship with some accuracy $ \alpha $.

\begin{equation}
	\begin{split}
		x &= \textrm{rand}([0, 1]) \\
		\mathrm{Expert}(\alpha) &= \begin{cases} 
			\mathrm{TrueReln}(X, Y),  & \textrm{if  } x <= \alpha \\
			\textrm{rand}(X \rightarrow Y, Y \leftarrow X, \textrm{None}) & \textrm{otherwise} \\
				\end{cases} \\
	\end{split}
\end{equation}

@@

\begin{equation}
	\begin{split}
	\mathrm{TrueReln}(X, Y) &= \begin{cases}
					X \rightarrow Y, & \textrm{if } X \rightarrow Y \in G \\
					Y \rightarrow X, & \textrm{if } Y \rightarrow X \in G \\
					\textrm{None}, & \textrm{otherwise }
				  \end{cases}
	\end{split}
\end{equation}

An important point to note here is that the effective accuracy of $
\mathrm{Expert} $ is higher than $ \alpha $ as even when giving a random
answer, there is a $ 1 $ in $ 3 $ chance that the answer is correct by chance.
Hence, for an $ \mathrm{Expert} $ with accuracy $ \alpha $, the effective
accuracy becomes:

@@

We analyzed the performance of \textsc{ExpertInLoop} algorithm by comparing it
with three other automated algorithms: PC, Hill-Climb Search, and Greedy
Equivalence Search (GES). For the analysis, we simulated data from known DAGs
and use the algorithms to recover the original DAG. We start by generating a
random DAG on $10$ nodes and use linear models with random effects to simulate
$ 500 $ samples from it. We simulate each variable $ Y_i $ in the DAG as:

\begin{equation}
	\begin{split}
		\beta &= \mathcal{N}(\bm{0}, \bm{I}) \\
		Y_i &= \bm{\beta} \mathrm{Pa}_G(Y_i) + \mathcal{N}(0, 1)
	\end{split}
\end{equation}

We perform the experiment for both varying densities of the DAG and accuracy of
$\mathrm{Expert}$. We also repeat each experiment $ 30 $ times and report the
mean and standard error of the results.

To compare the learned DAG to the original DAG, we use two metrics: Structural
Hamming Distance (SHD) and Structural Intervention Distance (SID)
\citep{Peters2015}. Unlike \textrm{ExpertInLoop}, the automated algorithms can
only recover the CPDAG, we consider all possible orientations of the CPDAG to
compute the range of SHD and SID values to show the best and worst case
results. The results of this analysis are shown in Figure~\ref{fig:shd_sid}.
For SHD, the performance of \textsc{ExpertInLoop} is comparable to the
automated algorithms for $ \alpha = 0.3 (\alpha_{\textrm{eff}} = 0.53 $, which
is equivalent to the expert getting the orientation correct $ 1 $ out of $ 2 $
case. For SID, the performance is comparable to the automated algorithms for $
\alpha = 0.5 (\alpha_{\textrm{eff}} = 0.67) $, which is equivalent to the
expert getting the orientation correct in $ 2 $ out of $ 3 $ cases. For denser
DAGs, the \textsc{ExpertInLoop} is able to perform better for lower $ \alpha $
values.
